---
title: "Reproduceable Research - Assignment 2"
author: "RHays"
date: "10/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r}
library(tidyverse)
library(readr)
library(ggplot2)
activity_data <- read_csv("activity.csv")
```
## 1. Total number of steps taken per day

```{r}
steps_per_day <- activity_data %>%
        filter(!is.na(steps)) %>% 
           group_by(date) %>% 
                summarize(steps = sum(steps)) %>%
  print
```

## 2. Histogram of the total number of steps taken each day

```{r}
hist(steps_per_day$steps, xlab = "Steps per Day", main = "Total Steps per Day", col = "green")
```
```{r}
mean_steps <- mean(steps_per_day$steps) %>% print()
median_steps <- median(steps_per_day$steps) %>% print()
```


Mean Steps per day 10766.19
Median Steps per day 10765


## 3. What is the average daily activity pattern?
```{r}
interval_steps <- activity_data %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))

ggplot(interval_steps, aes(x=interval, y=steps)) +
  geom_line(color = "blue")
```


Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Interval 835 has the the most steps at 206.17

```{r}
interval_steps[which.max(interval_steps$steps),]
```


#Calculate and report the total number of missing values in the dataset 
#Replace the NAs with the mean of exisiting Steps and Interval values
#Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity_data_na_populated <- activity_data
     nas <- sum(is.na(activity_data_na_populated$steps)) %>% print()
          avg_interval <- tapply(activity_data_na_populated$steps, activity_data_na_populated$interval, mean,                 na.rm=TRUE, simplify=TRUE)
          
activity_data_na_populated$steps[nas] <- avg_interval[as.character(activity_data_na_populated$interval[nas])]
View(activity_data_na_populated)
```

# Caluclate the total number of steps taken each day in the new data set 

```{r}
no_na_steps_per_day <- activity_data_na_populated %>%
        filter(!is.na(steps)) %>% 
           group_by(date) %>% 
                summarize(steps = sum(steps)) %>% print()
```
## Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. 
```{r}
hist(no_na_steps_per_day$steps, xlab = "Steps per Day", main = "Total Steps per Day", col = "blue")
```

## Calculate and report the mean and median total number of steps taken per day

```{r}
mean_steps <- mean(no_na_steps_per_day$steps) %>% print()
median_steps <- median(no_na_steps_per_day$steps) %>% print()

```

## Do these values differ from the estimates from the first part of the assignment?

The median calculations were the same for each data set. The mean was slightly lower in the second data set, by just 83.5 steps.

Calculations excluding NAs
[1] 10766.19
[1] 10765

Calculations Imputing NAs
[1] 10566.83
[1] 10682.5


```{r}
par(mfrow = c(1,2))
hist(steps_per_day$steps, xlab = "Steps per Day", main = "Total Steps per Day no NAs", col = "green")
hist(no_na_steps_per_day$steps, xlab = "Steps per Day", main = "Total Steps per Day NAs Imputed", col = "blue")
```

##What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
step_diff <- sum(activity_data$steps, na.rm=T) - sum(activity_data_na_populated$steps, na.rm=T)
print(step_diff)
```

There is a difference of -1.075472 in the total steps between the two data-sets.


## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable based on Weekday vs Weekend
2. Transform dayType variable into factor
```{r}
DayType <- function(date) {
  day <- weekdays(date)
  if (day %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))
      return ("Weekday")
  else if (day %in% c('Saturday', 'Sunday'))
      return ("Weekend")
  else
      stop ("Invalid Date Format.")
}
activity_data_na_populated$date <- as.Date(activity_data_na_populated$date)
activity_data_na_populated$day <- sapply(activity_data_na_populated$date, FUN = DayType)
unique(activity_data_na_populated$day)
```

```{r}
meanStepsByDay <- aggregate(steps ~ interval + day, activity_data_na_populated, mean)
ggplot(data = meanStepsByDay, aes(x = interval, y = steps)) + 
  geom_line() +
  facet_grid(day ~ .) +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```

Activity has an earlier start in the weekdays with a peak between 8AM and 9AM. On weekends, the activity starts later and does not have a peak between 8AM and 9AM, but does remain higher, on average, throughout weekends.
